### 2024-09-09 cognitive-bias.qmd
20240908 - regression to the mean

The regression fallacy is the failure to account for the fact that things tend to naturally fluctuate around their mean and that, after an extreme fluctuation, subsequent scores tend to regress (or reverse) to the mean.
An example of the regression fallacy is the so-called Sports Illustrated or Madden cover jinx curse.
The Sports Illustrated or Madden cover jinx curse is the urban legend that players who appear on the cover of Sports Illustrated (the magazine) or Madden (the video game) will perform poorly.
But, such a phenomenon can be more simply explained by regression to the mean (<https://www.psychologytoday.com/us/blog/what-the-luck/201610/the-sports-illustrated-cover-jinx>; archived at <https://perma.cc/CZM9-TVFN>).
When a player has a superb season, they likely benefited from some degree to good luck, and it is unlikely that they will repeat such a stellar season the following year.
Instead, they are likely—at least based on random fluctuation—to regress to their long-term mean.

>>>>

The regression fallacy is the failure to account for the fact that things tend to naturally fluctuate around their mean and that, after an extreme fluctuation, subsequent scores tend to regress (or reverse) to the mean.
Regression to the mean occurs because performance—both success and failure—is influenced by talent and by luck [@Kahneman2011].
And, unlike talent (or one's level one the construct of interest), luck is fickle.
Regression to the mean can explain why those with depression tend to improve over time without doing anything, and can explain why highly intelligent women tend to marry men who are less intelligent than they are [@Kahneman2011].
Despite peoples' tendency to ascribe such changes to causal phenomena, regression-related changes do not need a causal explanation—they are a mathematically inevitable consequence that arises from the fact that an outcome is influenced, to some degree, both by one's level on the construct and by error [@Kahneman2011].

An example of the regression fallacy is the so-called Sports Illustrated or Madden cover jinx curse.
The Sports Illustrated or Madden cover jinx curse is the urban legend that players who appear on the cover of Sports Illustrated (the magazine) or Madden (the video game) will perform poorly.
But, such a phenomenon can be more simply explained by regression to the mean (<https://www.psychologytoday.com/us/blog/what-the-luck/201610/the-sports-illustrated-cover-jinx>; archived at <https://perma.cc/CZM9-TVFN>) [@Kahneman2011].
When a player has a superb season, they likely benefited from some degree to good luck, and it is unlikely that they will repeat such a stellar season the following year.
Instead, they are likely—at least based on random fluctuation—to regress to their long-term mean.
